Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-01 LICS21/antitone.tex
Misc stuff
I don't even remember what changes this contains. Nothing major.

The following theorem does not prove completeness, but is suggestive in that direction:
\begin{theorem}
For any stopping time $T$ which is almost-surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
\end{theorem}

% !TEX root = ./probabilisticProgrammingMartingales.tex

\begin{proof}
If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
Otherwise, let $(t_n)_{n \geq 0}$ be defined recursively such that $t_0 = 0$, $t_{n+1} > t_n$ and $\mathbb P[T > t_{n+1}] \leq \frac 1 4 \mathbb P[T > t_n]$.
We will then define a supermartingale $(Y_n)_{n \geq 0}$ and nonrandom sequence $(y_n)_{n \geq 0}$ such that $Y_n = 0$ iff $T < n$, $Y_n = y_n$ iff $T \geq n$, and $y_n \geq 2^k$ for $n \geq t_k$.

The antitone function $\epsilon$ is defined piecewise and recursively in such a way as to force all the necessary constraints to hold:
\begin{align*}
    \text{for }x \in [0,1),\ &\epsilon(x) = 1 \\
    \text{for }x \in [2^k,2^{k+1}),\ &\epsilon(x) = \min \left (\epsilon(2^{k-1}), \frac{2^k}{t_{k+1}-t_k} \right ).
\end{align*}
This ensures that
\begin{itemize}
\item $\epsilon$ is weakly decreasing. 
\item $\epsilon$ is bounded below by 0.
\item For $x \geq 2^k$, $\epsilon(x) \leq \frac{2^k}{t_{k+1}-t_k}$. 
\lo{This would imply $\frac{2^k}{t_{k+1}-t_k} \geq \epsilon(2^{k-1})$. Why?} \akr{Corrected: it should have been min, not max.}
\end{itemize}

The sequence $(y_n)_{n \geq 0}$ is then defined recursively by
\begin{align*}
    y_0 = 2, \quad
    y_{n+1} = (y_n - \epsilon(y_n)) \frac{\mathbb P[T > n]}{\mathbb P[T > n+1]}.
\end{align*}

The fact $y_{t_k} \geq 2^{k+1}$ is proven by induction on $k$. For base case, $2 \geq 2$, as required. For the inductive case $k+1$, we first do another induction to prove that $y_n \geq 2^k$ for all $t_k \leq n \leq t_{k+1}$. The base case of this inner induction follows from the outer induction hypothesis a fortiori. Take the greatest $k$ such that $n > t_k$. By the induction hypothesis, $y_{t_k} \geq 2^k$.
\begin{align*}
    y_n & = y_{t_k} \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} - \sum_{m=t_k}^{n-1} \epsilon(y_m) \frac{\mathbb P[T > m+1]}{\mathbb P[T > n]} \\
    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \epsilon(y_m)) \\
    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \sum_{m=t_k}^{n-1} \frac{2^k}{t_{k+1}-t_k}) \\
    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - \frac{2^k (n - t_k)}{t_{k+1}-t_k}) \\
    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (y_{t_k} - 2^k) \\
    & \geq \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} (2^{k+1} - 2^k) \\
    & = \frac{\mathbb P[T > t_k]}{\mathbb P[T > n]} 2^k \\
    & \geq 2^k
\end{align*}
as required. Substituting $n = t_{k+1}$ and using the fact that $\frac{\mathbb P[T > t_k]}{\mathbb P[T > t_{k+1}]} > 4$, the same reasoning gives $y_{t_{k+1}} \geq 2^{k+2}$, as required.

Although the stronger induction hypothesis was necessary for the induction to work, the only reason this is needed is to prove that $y_n > 0$ for all $n$, which implies that $Y_n \geq 0$. 
The other condition for $(Y_n)$ to be an antitone-strict supermartingale is
\begin{align*}
    \expect{Y_{n+1} \mid \mathcal F_n} 
    &=  \expect{y_{n+1} \, {\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\
    &=  y_{n+1} \, \expect{{\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
    &=  y_{n+1} \, {\bf 1}_{\set{T \geq n}} \textstyle \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
    &=  (y_n - \epsilon(y_n)) \, {\bf 1}_{\set{T \geq n}} \\
    &= Y_n - \epsilon(y_n) \, {\bf 1}_{\set{T \geq n}}
\end{align*}
therefore $(Y_n)_n$ is an antitone strict supermartingale with respect to $(T,\epsilon)$.
\end{proof}

The assumption that $(\mathcal F_n)_{n \geq 0}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here. This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.

@@

\begin{example}[Higher-order recursion]
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined\footnote{Inspired by examples in \citep{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\quad\tif{n \leq 0}{s}{
f \; n \; (\varphi \; f \; s \; (n - 1))
\oplus_p
f \; (n+1) \; (\varphi \; f \; s \; n)
}
\end{array}
\]
\lo{QUESTION: Let $g : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function that terminates on all arguments.
Then $\Xi \, g$ is AST for $p \geq 1/2$.}
\end{example}

>>>>

The following theorem does not prove completeness, but is suggestive in that direction:
\begin{restatable}{theorem}{antitoneSeemsComplete}
For any stopping time $T$ which is almost-surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
\end{restatable}

@@

\begin{example}[Higher-order recursion]
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined\footnote{Inspired by examples in \citep{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\quad\tif{n \leq 0}{s}{
f \; n \; (\varphi \; f \; s \; (n - 1))
\oplus_p
f \; (n+1) \; (\varphi \; f \; s \; n)
}
\end{array}
\]
\lo{QUESTION: Let $g : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function that terminates on all arguments.
Then $\Xi \, g$ is AST for $p \geq 1/2$.}
\akr{In order to actually construct a ranking function, it will be necessary to assume a bound on the number of $\tY$-reduction steps in the reduction of $g \underline n \underline m$ too.}
\end{example}

### 2021-01-01 LICS21/ranking.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
Remove deprecated figures and comments from geometric distribution example.
Misc stuff
I don't even remember what changes this contains. Nothing major.

\begin{example}[Geometric distribution]
% Consider the term
% \begin{align*}
% &(\tY \lambda f, n. \\
% &\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
% &) \, \underline{0}
% \end{align*}
%\[
%\big(\tY \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
%\big) \, \underline{0}
%\]
Let
\[
\Theta := \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)}. 
\]
The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
\changed[akr]{Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.}
\lo{\Cref{fig:geometric distribution} is confusing: the arrow means either one-step reduction, or its reflexive transitive closure, or ``has the same syntactic shape''.}

\akr{The arrows all mean the one-step reduction $\to$, I just didn't intend for the values of $i$ and $r$ to be consistent between the terms. I would much rather have the diagram be complete (with all the reductions and all the terms).}

\lo{Perhaps you have in mind an informal, schematic interpretation, reading a term $C[i, r]$ as ``there exists a term $C[i, r]$, for appropriate $i, r \in \mathbb{R}$''.
One way to make this precise is to consider the graph obtained by quotienting the one-step reduction relation by an appropriate equivalence relation over terms (that specifies this scheme reading).}

\lo{\Cref{fig:geometric distribution 1} is less ambitious, but it is not confusing in that arrow does mean one-step reduction (also note $r'$), and it would be correct to say (cf.~the preceding blue-highlighted phrase): ``Despite its simplicity, $\mathit{Rch}(\tY \, \Theta \, \underline 0)$ contains all the terms in \Cref{fig:geometric distribution 1}, for each $i \geq 0$''.}

\lo{The graph can be simplified somewhat if we redefine general recursion using a fixpoint operator $\tY_{f, n}\Theta$ with redex rule as per \Cref{eq:letrec redex}: see \Cref{fig:geometric distribution 2}.
But (especially in view of the vast investment in the confluent trace semantics) it is perhaps too late to change things now?}

@@

    \draw [->] (9.west) to [bend left=45, "$i := i + 1$" {above,sloped}] (2.west);
    \draw [->] (6) to ["$r < 0.5$"] (7);
\end{tikzpicture}
\caption{The reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution}}
\end{figure}
\begin{figure}
\centering
\begin{tikzpicture}[node distance=0.5em, anchor=west]
    \node (1) at (0,-0) {$\tY\, \Theta\, \underline i$ \quad [provided $i = 0$]};
    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline i$};
    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})})\, \underline i$};
    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (6) at (0,-5) {$\tif{\underline r' < 0}{\underline i}{\tY\, \Theta\, (\underline i\, \underline{+ 1})}$};
    \node (7) at (2,-6) [text width=] {$\underline i$};
    \node (8) at (0,-7) {$\tY\, \Theta\, (\underline i\, \underline{+ 1})$};
    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
    \node (10) at (0,-9) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\, \Theta\, (n \underline{+ 1})}) z)\, \underline{i + 1}$};

    \draw [->] (1.south west) to [bend right=45] (2.north west);
    \draw [->] (2.south west) to [bend right=45] (3.north west);
    \draw [->] (3.south west) to [bend right=45] (4.north west);
    \draw [->] (4.south west) to [bend right=45] (5.north west);
    \draw [->] (5.south west) to [bend right=45] (6.north west);
    \draw [->] (6.south west) to [bend right=45] (8.north west);
    \draw [->] (8.south west) to [bend right=45] (9.north west);
    \draw [->] (9.south west) to [bend right=45] (10.north west);
    \draw [->] (6) to (7);
\end{tikzpicture}
\caption{Some reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution 1}}
\end{figure}

\begin{figure}
\centering
\begin{tikzpicture}[node distance=0.5em, anchor=west]
    \node (1) at (0,-0) {$\tY_{f, n}\Theta\, \underline i$};
    %\node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, \underline i$};
    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})})\, \underline i$};
    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
    \node (5) at (0,-4) {$\tif{\underline r\, \underline{- 0.5} < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
    \node (6) at (0,-5) {$\tif{\underline r' < 0}{\underline i}{\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})}$};
    \node (7) at (2,-6) [text width=] {$\underline i$};
    \node (8) at (0,-7) {$\tY_{f, n}\Theta\, (\underline i\, \underline{+ 1})$};
    \node (9) at (0,-8) {$\tY_{f, n}\Theta\, \underline{i + 1}$};
    %\node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, (\underline i\, \underline{+ 1})$};
    %\node (10) at (0,-9) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY_{f, n}\Theta\, (n \underline{+ 1})}) z)\, \underline{i + 1}$};

    \draw [->] (1.south west) to [bend right=45] (3.north west);
    %\draw [->] (2.south west) to [bend right=45] (3.north west);
    \draw [->] (3.south west) to [bend right=45] (4.north west);
    \draw [->] (4.south west) to [bend right=45] (5.north west);
    \draw [->] (5.south west) to [bend right=45] (6.north west);
    \draw [->] (6.south west) to [bend right=45] (8.north west);
    \draw [->] (8.south west) to [bend right=45] (9.north west);
    %\draw [->] (9.south west) to [bend right=45] (10.north west);
    \draw [->] (6) to (7);
\end{tikzpicture}
\caption{Some reachable terms of $\tY_{f, n}\Theta \, \underline 0$\label{fig:geometric distribution 2}}
\end{figure}

>>>>

\begin{example}[Geometric distribution]
% Consider the term
% \begin{align*}
% &(\tY \lambda f, n. \\
% &\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
% &) \, \underline{0}
% \end{align*}
%\[
%\big(\tY \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
%\big) \, \underline{0}
%\]
Let
\[
\Theta := \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)}. 
\]
The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$. The meanings of the substitutions and conditions that label some of the reductions in the diagram should be self-explanatory.

@@

    \draw [->] (9.west) to [bend left=45, "$i := i + 1$" {below,sloped}] (2.west);
    \draw [->] (6) to ["$r < 0.5$"] (7);
\end{tikzpicture}
\caption{The reachable terms of $\tY \, \Theta \, \underline 0$\label{fig:geometric distribution}}
\end{figure}
